Quantifying the error in forward modeling gravitational fields using tesseroids with adaptive discretization

### 2014-11-10 manuscript.tex
A bit of progress on methods
Need to better describe the adaptive discretization algorithm.
Duped by previous authors in dividing only when necessary. It was hidden
and obscured in the pseudo-code of the appendix. Focus on numerical
results.

\title[Tesseroid forward modeling with adaptive discretization]{
Forward modeling of gravitational fields using tesseroids with adaptive
discretization
}

\author[Uieda et al.]{
Leonardo Uieda$^{1,2}$,
Val\'eria C. F. Barbosa$^{2}$,
Vanderlei C. Oliveira Jr$^{2}$
\\
$^1$Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil.
$^2$Observat\'orio Nacional, Rio de Janeiro, Brazil.
}

@@

We propose an improvement over the adaptive discretization algorithm of
\citet{Li2011} to further increase the computational efficiency.

We will also quantify the error committed for the gravity gradients.

This is done in an attempt to establish an optimal minimum threshold for the
ratio between the distance to the computation point and the dimensions of the
tesseroid.

@@

>>>>

\title[Quantifying the error in tesseroid forward modeling]{
Quantifying the error in forward modeling gravitational fields using tesseroids
with adaptive discretization
}

\author[Uieda et al.]{
Leonardo Uieda$^{1,2}$,
Val\'eria C. F. Barbosa$^{2}$,
Vanderlei C. Oliveira Jr$^{2}$,
Carla Braitenberg$^{3}$
\\
$^1$Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil. \\
$^2$Observat\'orio Nacional, Rio de Janeiro, Brazil. \\
$^3$Department of Mathematics and Geosciences, University of Trieste, Trieste, Italy.
}

@@

We will attempt to quantify the error committed in the computation of the
gravity gradients using the GLQ with adaptive discretization.

We will establish an empirical relationship between the integration error and
the ratio between the distance to the computation point and the dimensions of
the tesseroid.

This will allow one to choose the appropriate amount of discretization to
achieve the desired accuracy.

@@

\citet{Grombein2013} provides Cartesian kernels for the tesseroid volume
internals that define the gravitational potential, gravitational attraction,
and Marussi tensor (EQUATIONS).
\begin{equation}
    V(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{1}{\ell} \kappa  dr' d\phi' d\lambda',
    \label{eq:tesspot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{\Delta\alpha}{\ell^3} \kappa dr' d\phi' d\lambda',
    \label{eq:tessgrav}
\end{equation}
\noindent
and
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        I_{\alpha\beta}
        dr' d\phi' d\lambda',
    \label{eq:tesstensor}
\end{equation}

\citet{Grombein2013} follows \citet{Heck2007} and solves the integrals by a
Taylor expansion around the geometric center of the tesseroid.

We will follow \citet{Asgharzadeh2007} and perform the numerical integration
using the Gauss-Legendre Quadrature (GLQ) instead.

The GLQ consists of approximating the integral by a weighted sum of the
integration kernel \citep{Hildebrand1987},
\begin{equation}
    \int\limits_a^b f(x) dx \approx
    \frac{b-a}{2}\sum\limits_{i=1}^N W_i f(x_i),
    \label{eq:glq1d}
\end{equation}

$N$ is the number of points used in the GLQ, i.e. the order of the quadrature.

The points $x_i$ where the kernel is evaluated are called the quadrature nodes.

The points are roots of the $N^{th}$ order Legendre polynomial $P_N(x)$.

The roots are $x = \pm 0.577350269$ for a second order polynomial ($P_2(x)$).

A root finder algorithm can be used to determine the roots for larger order
polynomials.

We refer the reader to \citet{Barrera-Figueroa2006} for an algorithm tailored
for Legendre polynomials in particular.

The second order roots, as well as the ones determined by a root finder
algorithm, will be within the range $[-1, 1]$.

Thus, the roots must be scaled to the integration limits $[a, b]$ by
\begin{equation}
    x^{scaled}_i = \frac{b - a}{2} x_i + \frac{b + a}{2}.
\end{equation}

The weights of the GLQ are given by \citep{Hildebrand1987},
\begin{equation}
    W_i = \frac{2}{(1 - x_i^2)(P^\prime_N(x_i))^2},
\end{equation}

The values of the $P_N(x)$ and its first derivative $P^\prime_N(x)$ can be
calculated with recursive relations (PUT FORMULA HERE?).

The Gauss-Legendre Quadrature for volume integrals, like equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
becomes \citep{Asgharzadeh2007} (EQUATION).

Applying equation (GLQ 3D) to equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
we see that $f(r_i, \phi_j, \lambda_k)$ is the effect of a point
mass located on the quadrature nodes.

It can be said that the GLQ integration approximates the volume integrals  by a
weighted sum of point mass effects.

\section{Adaptive discretization algorithm}

The quantity of, and therefore distance between, point masses regulates the
accuracy of the numerical integration \citep{Ku1977}.

\citet{Ku1977} determined an empirical criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.

Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.

One way to do this would be increase the order of the quadrature $N$ in all
three dimensions.

An alternative way would be to fix the quadrature order (e.g., $N=2$) and
instead divide the tesseroid in half along each dimension.

The gravitational effect of the larger tesseroid would be calculated as the sum
of the effects of the eight smaller tesseroids.

This division would effectively decrease the distance between nodes because of
the smaller tesseroids.

\citet{Li2011} developed an algorithm based on the tesseroid division approach.

In their algorithm, a tesseroid is divided in half along one of its dimensions
if the minimum distance to the computation point is smaller than a multiple of
the size of the tesseroid along that dimension.

The criterion check and (possible) subsequent division is repeated recursively
(i.e., to each of the smaller tesseroids and so on).

The recursion stops when none of the smaller tesseroids fail the criterion.

The advantage of this adaptive discretization is that the number of point
masses is only increased in regions of the tesseroid that are closer to the
computation point.

Alternatively, simply increasing the order of the GLQ would increase the number
of point masses evenly throughout the whole tesseroid.

The key controlling point of their algorithm is the ratio between
the smallest distance to the computation point and
the largest side length of the tesseroid.

The value of this ratio is used to decide whether a tesseroid will be divided
or not.

Thus, the value chosen for the minimum allowed value of this ratio determines
how many divisions are made.

Consequently, the value for the minimum distance-size ratio controls both the
accuracy of the integration and the computation time.

\citet{Li2011} provide a few examples of values for this ratio in the case of
the vertical component of gravitational attraction.

However, there has been no further investigation into
the relationship between values of the minimum distance-size
ratio and the computation error.

It is also not clear whether the same relationship holds for the gravitational
attraction as well as for the gradient tensor components.
